---
title: "Scenario Modelling"
author: " **Dr. Arun Mitra Peddireddy** <br> *JHAPSMCON 2026* <br> 4^th^ Annual Jharkhand State Conference of IAPSM"
format: revealjs
logo: aiims_deogarh_logo.jpg
include-after: |
  <script type="text/javascript">
    Reveal.on('ready', event => {
      if (event.indexh === 0) {
        document.querySelector("div.has-logo > img.slide-logo").style.display = "none";
      }
    });
    Reveal.addEventListener('slidechanged', (event) => {
      if (event.indexh === 0) {
        Reveal.configure({ slideNumber: null });
        document.querySelector("div.has-logo > img.slide-logo").style.display = "none";
      }
      if (event.indexh === 1) {
        Reveal.configure({ slideNumber: 'c' });
        document.querySelector("div.has-logo > img.slide-logo").style.display = null;
      }
    });
  </script>
---

## The Counterfactual

> *What would have happened without lockdown / masks / vaccines?*

We can never observe both as there is only one history.

**Models are the only way to answer counterfactual questions.**

Build a baseline. Build an alternative. 

**The comparison is the answer.**

::: {.notes}
Trajectory alone is not the result. Difference is the result.
:::

## Three Intervention Types

```{r}
#| echo: false
#| fig-width: 9.5
#| fig-height: 4.5
library(tidyverse); library(grid)

nodes_iv <- tibble(
  name = c("S", "E", "I", "R", "V"),
  x    = c(1, 3, 5, 7, 1),
  y    = c(0, 0, 0, 0, -1.4),
  fill = c("#1b9e77", "#e7298a", "#d95f02", "#7570b3", "#66a61e")
)
flow_edges <- tibble(
  from_x = c(1, 3, 5), to_x = c(3, 5, 7), y = 0,
  label  = c("β · I / N", "σ", "γ"),
  mid_x  = c(2, 4, 6)
)
buf <- 0.36

ggplot() +
  geom_segment(data = flow_edges,
               aes(x = from_x + buf, xend = to_x - buf, y = y, yend = y),
               arrow = arrow(length = unit(10, "pt"), type = "closed"),
               linewidth = 1.1, colour = "grey25") +
  geom_text(data = flow_edges, aes(x = mid_x, y = 0.4, label = label),
            size = 5.5) +
  annotate("segment",
           x = 1, xend = 1, y = -buf, yend = -1.4 + buf,
           arrow = arrow(length = unit(10, "pt"), type = "closed"),
           linewidth = 1.1, colour = "grey25") +
  annotate("text", x = 1.18, y = -0.7, label = "ν", size = 6, hjust = 0) +
  geom_point(data = nodes_iv, aes(x = x, y = y, colour = name), size = 24) +
  geom_text(data = nodes_iv, aes(x = x, y = y, label = name),
            colour = "white", fontface = "bold", size = 8) +
  annotate("segment", x = 1.4, xend = 2.6, y = 1.0, yend = 1.0,
           colour = "#1f6fb3", linewidth = 0.7) +
  annotate("segment", x = 1.4, xend = 1.4, y = 1.0, yend = 0.85,
           colour = "#1f6fb3", linewidth = 0.7) +
  annotate("segment", x = 2.6, xend = 2.6, y = 1.0, yend = 0.85,
           colour = "#1f6fb3", linewidth = 0.7) +
  annotate("text", x = 2, y = 1.35,
           label = "1) Reduce contacts\n(NPI: lower β)",
           size = 4.2, colour = "#1f6fb3", fontface = "bold") +
  annotate("text", x = 1.7, y = -0.7,
           label = "2) Vaccinate\n(move S → V)",
           size = 4.2, hjust = 0, colour = "#2f7c2a", fontface = "bold") +
  annotate("segment", x = 5.4, xend = 6.6, y = 1.0, yend = 1.0,
           colour = "#5d519a", linewidth = 0.7) +
  annotate("segment", x = 5.4, xend = 5.4, y = 1.0, yend = 0.85,
           colour = "#5d519a", linewidth = 0.7) +
  annotate("segment", x = 6.6, xend = 6.6, y = 1.0, yend = 0.85,
           colour = "#5d519a", linewidth = 0.7) +
  annotate("text", x = 6, y = 1.35,
           label = "3) Speed exit\n(treat / isolate)",
           size = 4.2, colour = "#5d519a", fontface = "bold") +
  scale_colour_manual(values = setNames(nodes_iv$fill, nodes_iv$name)) +
  coord_fixed(xlim = c(-0.3, 7.5), ylim = c(-2.0, 1.7), clip = "off") +
  theme_void() +
  theme(legend.position = "none", plot.margin = margin(8, 8, 8, 8))
```

## A Hypothetical COVID-Like Wave {.smaller}

A pathogen seeds an **age-structured population** (POLYMOD survey: 0–20, 20–40, 40+).
COVID-like timescales, with one-in-a-million initially infectious in each age band.

| Parameter             | Value     | Meaning                                          |
|-----------------------|----------:|--------------------------------------------------|
| `transmission_rate`   | 0.3       | per-contact infection rate                       |
| `infectiousness_rate` | 1/5       | rate of E → I  (5-day latent period)             |
| `recovery_rate`       | 1/7       | rate of I → R  (7-day infectious period)         |
| Population            | POLYMOD demography (3 age groups) | from `socialmixr::polymod`         |
| `time_end`            | 365 days  | one-year horizon                                 |

**The question:** how do three interventions compare against doing nothing?

## Building the Population {.smaller}

```{r}
#| code-line-numbers: "|1-2|4-8|10-13|15-22"
library(epidemics); library(socialmixr); library(tidyverse)

contact_data <- contact_matrix(                                # <1>
  socialmixr::polymod,
  age_limits = c(0, 20, 40),
  symmetric  = TRUE
)

contact_matrix    <- t(contact_data[["matrix"]])               # <2>
demography_vector <- contact_data[["demography"]][["population"]]
names(demography_vector) <- rownames(contact_matrix)

initial_i <- 1e-6                                              # <3>
init_row  <- c(S = 1 - initial_i, E = 0, I = initial_i, R = 0, V = 0)
initial_conditions <- rbind(init_row, init_row, init_row)
rownames(initial_conditions) <- rownames(contact_matrix)

uk_population <- population(                                   # <4>
  name               = "Population",
  contact_matrix     = contact_matrix,
  demography_vector  = demography_vector,
  initial_conditions = initial_conditions
)
```

1. POLYMOD survey of who-meets-whom — used to build a contact matrix per day, per age group.
2. Three age groups → a 3 × 3 contact matrix C, plus a demography vector with one population count per group.
3. Initial conditions as **proportions** per age group (epidemics expects rows summing to 1).
4. The `population` object bundles structure + demography + starting state.

## A Helper To Run Scenarios {.smaller}

```{r}
#| code-line-numbers: "|3-7|9-12"
run_scenario <- function(npi = NULL, vac = NULL, t_end = 365) {

  args <- list(                                                # <1>
    population          = uk_population,
    transmission_rate   = 0.3,                # COVID-like
    infectiousness_rate = 1 / 5,
    recovery_rate       = 1 / 7,
    time_end            = t_end,
    increment           = 1.0
  )

  if (!is.null(npi)) args$intervention <- list(contacts = npi) # <2>
  if (!is.null(vac)) args$vaccination  <- vac

  do.call(model_default, args)
}
```

1. **Baseline arguments** — same for every scenario.
2. **Optional levers** — pass an `intervention()` object for an NPI, a `vaccination()` object for a campaign, or both. No arguments → counterfactual.

## Scenario 0 — Counterfactual (Do Nothing) {.smaller}

```{r}
#| message: false
counterfactual <- run_scenario()
```

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 3.6
counterfactual |>
  filter(compartment %in% c("susceptible", "infectious", "recovered")) |>
  group_by(time, compartment) |>
  summarise(value = sum(value), .groups = "drop") |>
  mutate(compartment = factor(compartment,
                              levels = c("susceptible", "infectious", "recovered"),
                              labels = c("S", "I", "R"))) |>
  ggplot(aes(time, value, colour = compartment)) +
  geom_line(linewidth = 1.1) +
  scale_colour_manual(values = c(S = "#1b9e77", I = "#d95f02", R = "#7570b3")) +
  scale_y_continuous(labels = scales::comma) +
  labs(x = "Day", y = "People", colour = NULL,
       title = "Counterfactual — let it rip") +
  theme_minimal(base_size = 14) + theme(legend.position = "top")
```

## Scenario A — 60-Day Lockdown (NPI) {.smaller}

```{r}
#| message: false
#| code-line-numbers: "|4|5-6|7"
lockdown_60 <- intervention(
  name       = "Lockdown",
  type       = "contacts",
  reduction  = matrix(c(0.5, 0.5, 0.3)),    # <1>
  time_begin = 30,
  time_end   = 90                            # <2>
)

scenario_npi <- run_scenario(npi = lockdown_60)
```

1. Reduction **per age group**: kids 50%, working-age 50%, older 30%. Older population is already isolating, so a smaller marginal change.
2. The intervention is on for 60 days; β returns to baseline on day 91.

## Scenario A — Plot

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 3.4
total_I <- function(out) {
  out |> filter(compartment == "infectious") |>
    group_by(time) |> summarise(I = sum(value), .groups = "drop")
}
bind_rows(
  total_I(counterfactual) |> mutate(scenario = "Counterfactual"),
  total_I(scenario_npi)   |> mutate(scenario = "NPI (days 30–90)")
) |>
  ggplot(aes(time, I, colour = scenario)) +
  annotate("rect", xmin = 30, xmax = 90, ymin = 0, ymax = Inf,
           fill = "#1f6fb3", alpha = 0.12) +
  geom_line(linewidth = 1.1) +
  scale_colour_manual(values = c("Counterfactual" = "grey60",
                                 "NPI (days 30–90)" = "#1f6fb3")) +
  scale_y_continuous(labels = scales::comma) +
  labs(x = "Day", y = "Infectious", colour = NULL,
       title = "Lockdown delays the peak — but the wave returns") +
  theme_minimal(base_size = 14) + theme(legend.position = "top")
```

## Scenario B — Vaccination Campaign {.smaller}

```{r}
#| message: false
#| code-line-numbers: "|3|4|5"
campaign <- vaccination(
  name        = "Campaign",
  time_begin  = matrix(60,    nrow = 3),     # <1>
  time_end    = matrix(360,   nrow = 3),     # <2>
  nu          = matrix(0.005, nrow = 3)      # <3>
)

scenario_vac <- run_scenario(vac = campaign)
```

1. **Start day per age group.** Matrix shape n_age × n_doses (here 3 × 1). Same start day for everyone here.
2. **End day per age group.** A 300-day campaign window.
3. **Daily vaccination rate** — fraction of remaining S moved to V per day. 0.5%/day.

## Scenario B — Plot

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 3.4
bind_rows(
  total_I(counterfactual) |> mutate(scenario = "Counterfactual"),
  total_I(scenario_vac)   |> mutate(scenario = "Vaccination (from day 60)")
) |>
  ggplot(aes(time, I, colour = scenario)) +
  geom_vline(xintercept = 60, linetype = "dashed", colour = "#2f7c2a") +
  geom_line(linewidth = 1.1) +
  scale_colour_manual(values = c("Counterfactual" = "grey60",
                                 "Vaccination (from day 60)" = "#2f7c2a")) +
  scale_y_continuous(labels = scales::comma) +
  labs(x = "Day", y = "Infectious", colour = NULL,
       title = "Vaccination shrinks the wave — but timing matters") +
  theme_minimal(base_size = 14) + theme(legend.position = "top")
```

## Scenario C — Combined (NPI + Vaccination) {.smaller}

```{r}
#| message: false
scenario_combined <- run_scenario(npi = lockdown_60, vac = campaign)
```

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 3.4
bind_rows(
  total_I(counterfactual)    |> mutate(scenario = "Counterfactual"),
  total_I(scenario_combined) |> mutate(scenario = "Combined (NPI + Vaccination)")
) |>
  ggplot(aes(time, I, colour = scenario)) +
  annotate("rect", xmin = 30, xmax = 90, ymin = 0, ymax = Inf,
           fill = "#1f6fb3", alpha = 0.12) +
  geom_vline(xintercept = 60, linetype = "dashed", colour = "#2f7c2a") +
  geom_line(linewidth = 1.1) +
  scale_colour_manual(values = c("Counterfactual" = "grey60",
                                 "Combined (NPI + Vaccination)" = "#5d519a")) +
  scale_y_continuous(labels = scales::comma) +
  labs(x = "Day", y = "Infectious", colour = NULL,
       title = "Lockdown buys the time vaccination needs") +
  theme_minimal(base_size = 14) + theme(legend.position = "top")
```

## All Four on One Plot

```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 5
all_scenarios <- bind_rows(
  total_I(counterfactual)    |> mutate(scenario = "0. Counterfactual"),
  total_I(scenario_npi)      |> mutate(scenario = "1. NPI only"),
  total_I(scenario_vac)      |> mutate(scenario = "2. Vaccination only"),
  total_I(scenario_combined) |> mutate(scenario = "3. Combined")
)

ggplot(all_scenarios, aes(time, I, colour = scenario)) +
  geom_line(linewidth = 1.2) +
  scale_colour_manual(values = c("0. Counterfactual"   = "grey50",
                                 "1. NPI only"         = "#1f6fb3",
                                 "2. Vaccination only" = "#2f7c2a",
                                 "3. Combined"         = "#5d519a")) +
  scale_y_continuous(labels = scales::comma) +
  labs(x = "Day", y = "Infectious", colour = NULL,
       title = "COVID-like wave — four futures from one model") +
  theme_minimal(base_size = 14) + theme(legend.position = "top")
```

## Outcomes

```{r}
#| echo: false
totals <- function(out) {
  out |> filter(compartment %in% c("infectious", "recovered", "vaccinated")) |>
    group_by(time, compartment) |> summarise(value = sum(value), .groups = "drop") |>
    pivot_wider(names_from = compartment, values_from = value)
}
counter_R <- max(totals(counterfactual)$recovered)

bind_rows(
  totals(counterfactual)    |> mutate(scenario = "0. Counterfactual"),
  totals(scenario_npi)      |> mutate(scenario = "1. NPI only"),
  totals(scenario_vac)      |> mutate(scenario = "2. Vaccination only"),
  totals(scenario_combined) |> mutate(scenario = "3. Combined")
) |>
  group_by(scenario) |>
  summarise(
    `Peak day`        = time[which.max(infectious)],
    `Peak infectious` = scales::comma(round(max(infectious))),
    `Final size`      = scales::comma(round(max(recovered))),
    `Vaccinated`      = scales::comma(round(max(vaccinated))),
    `Cases averted`   = scales::comma(round(counter_R - max(recovered))),
    .groups = "drop"
  ) |>
  knitr::kable()
```

::: fragment
**NPIs buy time.**
:::
::: fragment
**Vaccination prevents.**
:::
::: fragment
**Combined wins.**
:::

## Hands-on {.smaller}

Open `scripts/scenario_explorer.R`. Setup loads the population + helpers; pick **one** task per pair.

| Pair | Task                                                                          |
|------|-------------------------------------------------------------------------------|
| **A** | Weak lockdown — change `reduction` to `c(0.3, 0.3, 0.2)`. Does the wave still flatten? |
| **B** | Long lockdown — extend to days 30–150. Smaller rebound?                       |
| **C** | Late lockdown — start day 60 instead of day 30. How does peak day shift?      |
| **D** | Earlier vaccination — start day 30 instead of day 60. Final size change?      |
| **E** | Slow vaccination — `nu = 0.001` instead of 0.005. Still useful?               |
| **F** | Vaccinate only — drop the NPI, but double `nu` to 0.01. Trade-off?            |

For your tweak: **predict first**, then run, then write down peak day / peak I / final size.

## Reveal — Pair Results {.smaller}

```{r}
#| echo: false
#| message: false
#| fig-width: 11
#| fig-height: 5

pairs <- tribble(
  ~pair, ~label,                       ~reduction,                   ~npi_start, ~npi_end, ~nu,    ~vac_start, ~vac_end,
  "A",   "Weak NPI (0.3, 0.3, 0.2)",   c(0.3, 0.3, 0.2),             30,         90,       0,      0,          0,
  "B",   "Long NPI (30–150)",          c(0.5, 0.5, 0.3),             30,         150,      0,      0,          0,
  "C",   "Late NPI (60–120)",          c(0.5, 0.5, 0.3),             60,         120,      0,      0,          0,
  "D",   "Early vac (day 30)",         NULL,                          0,          0,       0.005,  30,         360,
  "E",   "Slow vac (0.001/day)",       NULL,                          0,          0,       0.001,  60,         360,
  "F",   "Vac only, faster (0.01)",    NULL,                          0,          0,       0.010,  30,         360
)

build_npi <- function(reduction, t0, t1) {
  if (is.null(reduction) || all(reduction == 0)) return(NULL)
  intervention(name = "NPI", type = "contacts",
               reduction = matrix(reduction),
               time_begin = t0, time_end = t1)
}
build_vac <- function(nu, t0, t1) {
  if (is.null(nu) || nu == 0) return(NULL)
  vaccination(name = "Vac",
              time_begin = matrix(t0,  nrow = 3),
              time_end   = matrix(t1,  nrow = 3),
              nu         = matrix(nu,  nrow = 3))
}

pair_runs <- pairs |>
  mutate(
    sim = pmap(list(reduction, npi_start, npi_end, nu, vac_start, vac_end),
               \(r, ts, te, n, vs, ve) run_scenario(npi = build_npi(r, ts, te),
                                                    vac = build_vac(n, vs, ve))),
    panel = paste0("Pair ", pair, " — ", label)
  )

pair_totals <- pair_runs |>
  mutate(I_curve = map(sim, total_I)) |>
  select(panel, I_curve) |>
  unnest(I_curve)

ggplot(pair_totals, aes(time, I)) +
  geom_line(data = total_I(counterfactual),
            aes(time, I), colour = "grey70",
            linewidth = 0.7, linetype = "dashed") +
  geom_line(linewidth = 1.1, colour = "#5d519a") +
  facet_wrap(~ panel, ncol = 3) +
  scale_y_continuous(labels = scales::comma) +
  labs(x = "Day", y = "Infectious",
       title = "Each pair's tweak vs the counterfactual (dashed)") +
  theme_minimal(base_size = 12) + theme(strip.text = element_text(face = "bold"))
```

## Closing 

> A trajectory tells you what *could* happen.<br>
> A counterfactual tells you **what to do**.

<br>

:::{.fragment}
### Questions?
:::